{\bf Affine.m} -- {\it Mathematica} package for computations in representation theory of finite-dimensional and affine Lie algebras

### 2011-06-11 doc/paper.tex
Sum of modules

The structure of affine Lie algebras allows to extend computational algorithms created for finite-dimensional Lie algebras  \cite{Fuchs:1996dd}, \cite{gannon2001algorithms}, \cite{kass1990ala}. The book \cite{kass1990ala} with the tables of multiplicities and other computed characteristics of algebras and representations was published in 1990. But we are not aware of software packages for popular computer algebra systems which can be used to extend these results.
We address this issue and present {\bf Affine.m} -- {\it Mathematica} package for computations in representation theory of affine and finite-dimensional Lie algebras.  We describe the features and limitations of the package in present paper and in the manual \cite{affinemanual}. In present paper we also provide representation-theoretical background of implemented algorithms and present some examples of computations relevant to physics.

The paper is started with an overview of Lie algebras and their representation theory (Sec. \ref{sec:theor-backgr}). Then we describe datastructures of {\bf Affine.m} used to present different objects related to Lie algebras and their representations (Sec. \ref{sec:core-datastructures} and used in implemented algorithms (Sec. \ref{sec:comp-algor}). Next section consists of physically interesting examples (Sec. \ref{sec:examples}). The paper is concluded with the discussion of possible extensions and refinements (Sec. \ref{sec:conclusion}).

@@

@@

Lie algebra $\gf$ is a vector space with bilinear operation $[\cdot,\cdot]:\gf\otimes\gf\to \gf$, which is called commutator. If we choose the basis $X_{i}$ in $\gf$ we can specify commutation relations by the structure constants $C_{ijk}$:
\begin{equation}
  \label{eq:1}
  [X^{i},X^{j}]=\sum_{k} C^{ij}_{k} X^{k}
\end{equation}
Lie algebra is simple if it contains no non-trivial ideals. Semisimple Lie algebra is a direct sum of simple Lie algebras.

@@

Killing form on $\gf$ gives a non-degenerate bilinear form $(\cdot,\cdot)$ on $\hfg$ which can be used to identify $\hfg$ with the subspace of the dual space $\hfg^{*}$ of linear functionals on $\hfg$. Weights are the elements of $\hfg^{*}$ and are denoted by Greek letters $\mu,\nu, \omega, \lambda\dots$

@@

Loop algebra $L\gf=\gf\otimes \mathbb{C}[t,t^{-1}]$, corresponding to semisimple Lie algebra $\gf$ has commutation relations
\begin{equation}
  \label{eq:6}
  [X^{i}t^{n},X^{j}t^{m}]=t^{n_+m}\sum_{k}C^{ij}_{k}X^{k}
\end{equation}
Central extension leads to the appearance of additional term
\begin{equation}
  \label{eq:7}
   [X^{i}t^{n}+\alpha c,X^{j}t^{m}+\beta c]=t^{n+m}\sum_{k}C^{ij}_{k}X^{k}+(X^{i},X^{j})n\delta_{n+m,0}c
\end{equation}
This algebra $\hat\gf=\gf\otimes\mathbb{C}[t,t^{-1}]\oplus\mathbb{C}c$ is called {\it affine Lie algebra} \cite{kac1990idl}, \cite{wakimoto2001idl,wakimoto2001lectures}, \cite{kass1990ala}.

@@

\subsection{Weights and roots}
\label{sec:weights-roots}

@@

Then $\gf$-module is a vector space $V$ together with a bilinear map $\gf \times V\to V$ such that
\begin{equation}
  \label{eq:2}
  [x,y]\cdot v = x\cdot(y\cdot v) - y\cdot(x\cdot v), \quad \mbox{for}\; x,y\in \gf, v\in V
\end{equation}
Representation of algebra $\gf$ on a vector space $V$ is the homomorphism $\gf\to gl(V)$.

@@

The structure of module is determined by the set of weights since the action of generators $E^{\alpha}$ on a weight vectors is
\begin{equation}
  \label{eq:5}
  E^{\alpha}\cdot v_{\lambda} \propto v_{\lambda+\alpha}
\end{equation}
It can be encoded by the formal character of module
\begin{equation}
  \label{eq:10}
  \mathrm{ch}V=\sum_{\lambda}m_{\lambda} e^{\lambda}
\end{equation}
Character  $\mathrm{ch}V\in \mathcal{E}$ is an element of algebra $\mathcal{E}$ generated by formal exponents of weights.
Character can be specialized by taking its value on some element $\xi$ of $\hf$:

Roots are weights of adjoint representation of $\gf$ on itself. They encode the commutation relations of algebra in the following way. Denote by $\Delta$ the set of roots. For each $\alpha\in \Delta$ there exist a root $-\alpha\in \Delta$ and generators $E^{\alpha}, E^{-\alpha}$ such that
\begin{align}
  \label{eq:4}
  &  [H^{i},E^{\alpha}]=\alpha^{i}E^{\alpha} \\
  &\left[E^{\alpha},E^{\beta}\right]=
  \begin{cases}
    N_{\alpha,\beta} E^{\alpha+\beta}, & \mbox{if}\; \alpha+\beta\in \Delta\\
    \frac{2}{(\alpha,\alpha)} \sum_{i}\alpha^{i} H^{i},&  \mbox{if}\; \alpha=-\beta\\
    0,&\mbox{otherwise}
  \end{cases}
\end{align}
Highest weight of adjoint representation.

@@

A positive root is simple if it cannot be written as the sum of positive roots. The set of simple roots $\left\{\alpha_{i}\right\}$ is a basis of $\hfg$ and each root can be written as $\alpha=\sum_{i}n_{i}\alpha_{i}$ with all $n_{i}$ non-negative or non-positive. In case of finite-dimensional Lie algebra $\gf$ simple roots are numbered from 1 to rank of the algebra $i=1,\dots,r,\quad r=\mathrm{rank}(\gf)$. By numbering simple roots with an index $i$ we introduce an ordering on the root system $\Delta$. Highest root with respect to this ordering is denoted by  $\theta=\sum_{i=1,\dots,r} a_i \alpha_i$, coefficients $a_i$ are called {\it marks}. {\it Comarcs} are equal to $a_i^{\vee}=\frac{(\alpha_i,\alpha_i)}{2} a_i$.

Although the full set of roots $\Delta$ is infinite for affine Lie algebra $\hat\gf$ the number of simple roots is finite and its elements are denoted by $\alpha_{0},\dots \alpha_{r}$ where $r=\mathrm{rank}(\gf)$. The roots $\alpha_1,\dots, \alpha_r$ are the roots of the underlying finite-dimensional Lie algebra $\go$. The root $\alpha_0=\delta-\theta$ is the difference of {\it imaginary root} $\delta$ and $\theta$ -- highest root of the algebra $\go$.

>>>>

The structure of affine Lie algebras allows to extend computational algorithms created for finite-dimensional Lie algebras  \cite{Fuchs:1996dd}, \cite{gannon2001algorithms}, \cite{kass1990ala}. The book \cite{kass1990ala} with the tables of multiplicities and other computed characteristics of affine Lie algebras and representations was published in 1990. But we are not aware of software packages for popular computer algebra systems which can be used to extend these results.
We address this issue and present {\bf Affine.m} -- {\it Mathematica} package for computations in representation theory of affine and finite-dimensional Lie algebras.  We describe the features and limitations of the package in present paper and in the manual \cite{affinemanual}. In present paper we also provide representation-theoretical background of implemented algorithms and present some examples of computations relevant to physics.

The paper is started with an overview of Lie algebras and their representation theory (Sec. \ref{sec:theor-backgr}). Then we describe datastructures of {\bf Affine.m} used to present different objects related to Lie algebras and representations (Sec. \ref{sec:core-datastructures}) and discuss implemented algorithms (Sec. \ref{sec:comp-algor}). Next section consists of physically interesting examples (Sec. \ref{sec:examples}). The paper is concluded with the discussion of possible extensions and refinements (Sec. \ref{sec:conclusion}).

@@

In this section we remind necessary definitions and present formulae used in computations.

@@

{\it Lie algebra} $\gf$ is a vector space with bilinear operation $[\cdot,\cdot]:\gf\otimes\gf\to \gf$, which is called {\it commutator}. If we choose the basis $X_{i}$ in $\gf$ we can specify commutation relations by the {\it structure constants} $C_{ijk}$:
\begin{equation}
  \label{eq:1}
  [X^{i},X^{j}]=\sum_{k} C^{ij}_{k} X^{k}
\end{equation}
Lie algebra is simple if it contains no non-trivial ideals. Semisimple Lie algebra is a direct sum of simple Lie algebras.

@@

Killing form on $\gf$ gives a non-degenerate bilinear form $(\cdot,\cdot)$ on $\hfg$ which can be used to identify $\hfg$ with the subspace of the dual space $\hfg^{*}$ of linear functionals on $\hfg$. {\it Weights} are the elements of $\hfg^{*}$ and are denoted by Greek letters $\mu,\nu, \omega, \lambda\dots$

@@

{\it Loop algebra} $L\gf=\gf\otimes \mathbb{C}[t,t^{-1}]$, corresponding to semisimple Lie algebra $\gf$, has commutation relations
\begin{equation}
  \label{eq:6}
  [X^{i}t^{n},X^{j}t^{m}]=t^{n_+m}\sum_{k}C^{ij}_{k}X^{k}
\end{equation}
Central extension leads to the appearance of additional term
\begin{equation}
  \label{eq:7}
   [X^{i}t^{n}+\alpha c,X^{j}t^{m}+\beta c]=t^{n+m}\sum_{k}C^{ij}_{k}X^{k}+(X^{i},X^{j})n\delta_{n+m,0}c
\end{equation}
This algebra $\hat\gf=\gf\otimes\mathbb{C}[t,t^{-1}]\oplus\mathbb{C}c$ is called {\it affine Lie algebra} \cite{kac1990idl}, \cite{wakimoto2001idl,wakimoto2001lectures}, \cite{kass1990ala}.

@@

\subsection{Modules, weights and roots}
\label{sec:weights-roots}

@@

Then $\gf$-module is a vector space $V$ together with a bilinear map $\gf \times V\to V$ such that
\begin{equation}
  \label{eq:2}
  [x,y]\cdot v = x\cdot(y\cdot v) - y\cdot(x\cdot v), \quad \mbox{for}\; x,y\in \gf, v\in V
\end{equation}
Representation of algebra $\gf$ on a vector space $V$ is the homomorphism $\gf\to gl(V)$ from $\gf$ to Lie algebra of endomorphisms on vector space $V$ with the commutator as the bracket.

@@

The structure of module is determined by the set of weights since the action of generators $E^{\alpha}$ on a weight vectors is
\begin{equation}
  \label{eq:5}
  E^{\alpha}\cdot v_{\lambda} \propto v_{\lambda+\alpha}
\end{equation}
Module structure can be encoded by the formal character of module
\begin{equation}
  \label{eq:10}
  \mathrm{ch}V=\sum_{\lambda}m_{\lambda} e^{\lambda}
\end{equation}
Character  $\mathrm{ch}V\in \mathcal{E}$ is an element of algebra $\mathcal{E}$ generated by formal exponents of weights.
Character can be specialized by taking its value on some element $\xi$ of $\hf$:

Lie algebra is its own module. The action of generators is called adjoint and it is given by the bracket $ad_{X} Y=[X,Y]$. 
{\it Roots} are weights of adjoint representation of $\gf$. They encode the commutation relations of algebra in the following way. Denote by $\Delta$ the set of roots. For each $\alpha\in \Delta$ there exist a root $-\alpha\in \Delta$ and generators $E^{\alpha}, E^{-\alpha}$ such that
\begin{align}
  \label{eq:4}
  &  [H^{i},E^{\alpha}]=\alpha^{i}E^{\alpha} \\
  &\left[E^{\alpha},E^{\beta}\right]=
  \begin{cases}
    N_{\alpha,\beta} E^{\alpha+\beta}, & \mbox{if}\; \alpha+\beta\in \Delta\\
    \frac{2}{(\alpha,\alpha)} \sum_{i}\alpha^{i} H^{i},&  \mbox{if}\; \alpha=-\beta\\
    0,&\mbox{otherwise}
  \end{cases}
\end{align}

@@

A positive root is {\it simple} if it cannot be written as the sum of positive roots. The set of simple roots $\Phi=\left\{\alpha_{i}\right\}$ is a basis of $\hfg$ and each root can be written as $\alpha=\sum_{i}n_{i}\alpha_{i}$ with all $n_{i}$ non-negative or non-positive. In case of finite-dimensional Lie algebra $\gf$ simple roots are numbered from 1 to rank of the algebra $i=1,\dots,r,\quad r=\mathrm{rank}(\gf)$. By numbering simple roots with an index $i$ we introduce an ordering on the root system $\Delta$. Highest root with respect to this ordering is denoted by  $\theta=\sum_{i=1,\dots,r} a_i \alpha_i$, coefficients $a_i$ are called {\it marks}. It is also highest weight (See section \ref{sec:high-weight-modul}) of adjoint module. {\it Comarcs} are equal to $a_i^{\vee}=\frac{(\alpha_i,\alpha_i)}{2} a_i$.

Although the full set of roots $\Delta$ is infinite for affine Lie algebra $\hat\gf$ the set of simple roots $\Phi$ is finite and its elements are denoted by $\alpha_{0},\dots \alpha_{r}$ where $r=\mathrm{rank}(\gf)$. The roots $\alpha_1,\dots, \alpha_r$ are the roots of the underlying finite-dimensional Lie algebra $\go$. The root $\alpha_0=\delta-\theta$ is the difference of {\it imaginary root} $\delta$ and $\theta$ -- highest root of the algebra $\go$.
